Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-15 doc/manuscript_hkk_drought_sensitivity.Rmd
replace fig 1 #10  to reflect #26
intro refs added

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. 
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Finally, the microenvironments that individual trees experience influence growth responses to drought. <!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted [e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021] or acclimatized [@refs].
Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems [@refs] and thereby potential to access more reliable water sources [@refs].
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

![***Figure `r fig_droughts`. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values at four different time scales for each month in 2010 and 2015. A _k_-month SPEI integrates over _k_ months ending in the month current month. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1.png)

@@

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event [@REF]. 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with two or more consecutive months with 1-month SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.

@@

<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings [n=4 at HKK; @vlam_temperature_2014] are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the two events; it's growth nearly doubled in the dry-season drought but was severely reduced in the wet-season drought (Fig. S5).
More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to directly and interactively influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought, in alignment with recent global analyses (@liu_drought_2022).

>>>>

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought, or hot/dry conditions anomalously departing from climatic means during any time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @brando_droughts_2019], droughts with different characteristics may in fact have very different effects on tree growth [@gao_dynamic_2018].
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
In tropical forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically depending on the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. 
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that tropical tree growth <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly sensitive to dry season precipitation [@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tropical tree species <!--in tropical dry forests -->respond to different types of droughts, including those occurring at different times relative to regular seasonal cycles<!-- of wet and dry seasons-->. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species adaptations to cope with drought vary widely in type and strength, often leading to classification of species as "drought tolerant" or "drought sensitive" [@guillemot_small_2022; @vico_how_2017a; @oliveira_linking_2021].
In tropical seasonal forests in particular, trees have evolved to maximise growth, reproduction, and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency, leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021].
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid foliar water loss and have acquisitive strategies during leaf-on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf water potential at turgor loss point ($\pi_{tlp}$, hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), and *large vessels (hydraulic efficiency)*<!--but this may put them at risk of embolism--> [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity, especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls.
For example, some species regulate negative $\pi_{tlp}$'s, leaf deciduousness, and root depth based on environmental cues, leading to intraspecific or interannual variation and potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; @hulshof_variation_2010].

<!--PARAGRAPH ON HOW MICROENVIRONMENT SHAPES DROUGHT RESPONSES-->
Finally, the microenvironments that individual trees experience influence growth responses to drought. <!--leading to varied atmospheric and hydrological stresses with distinct ecological consequences.-->
Across horizontal axes, microenvironments defined by topographic and edaphic variation shape tree drought responses. 
While habitats with relatively moist soils (e.g., low-lying topographic positions) are clearly more hospitable during drought, trees in these microenvironments may be more vulnerable if less drought-adapted (e.g., less negative $\pi_{tlp}$, @kunert_leaf_2021) or acclimatized [@refs].
Indeed, there are observations of trees near streams undergoing greater growth declines [@mcgregor_tree_2021] and increases in mortality [@zuleta_droughtinduced_2017] during drought. 
Across vertical axes, trees of different sizes have their crowns and root systems in sometimes dramatically divergent microenvironments.
Aboveground, tree crowns in canopy or emergent positions are exposed to higher evaporative demand and thermal stress driven by higher solar radiation, wind speed, and vapor pressure deficit [@vinod_thermal_2023], and such exposure makes trees more vulnerable to drought [@scharnweber_confessions_2019; @ma_tree_2023].
<!--In contrast, understory trees may be rarely exposed to direct solar radiation and extreme evaporative demand.-->
These aboveground gradients in crown exposure are at least partially offset by the tendency for large trees to have larger and often deeper root systems and greater stem water storage [@fernández-de-uña_role_2023] and thereby potential to access more reliable water sources [@stahl_depth_2013].
However, trees with deeper roots that rely on regular access to deep water may be more vulnerable during severe droughts when those sources are depleted [@chitra-tarak_hydraulicallyvulnerable_2021]. 
While the overall tendency is that larger trees suffer greater growth declines during drought [@bennett_larger_2015], it remains unclear whether this is driven primarily by crown exposure, by greater height itself makes trees more vulnerable to drought based on the physics of hydraulic flow through a porous medium <!--, as described by Darcy's law -->[@mcdowell_darcy_2015; @fernández-de-uña_role_inpress], or by trait differences across vertical axis [@vinod_thermal_2023].
Thus, horizontal and vertical microenvironmental gradients modify tree drought responses, likely in interaction with drought characteristics and species traits.

@@

![***Figure `r fig_droughts`. Climatic characteristics of drought years and average years*** in the Huai Kha Khaeng ForestGEO plot. For four daily climate variables dry days, precipitation, maximum temperature and VPD, panel a) represents rolling means for 30 days around the focal day for that year. Grey line represents the mean of this rolling mean from 2008 to 2019 and ribbons represent standard deviation around this mean. Panel b) represents anomalies of this rolling mean from the values across the timeseries. Values, therefore represents the number of standard deviations from the mean on a given day. c) shows Standardised Precipitation Evapotranspiration Index (SPEI) values at four different time scales for each month in 2010 and 2015. A _k_-month SPEI integrates over _k_ months ending in the month current month. Across all panels, light blue background represents the average wet season, from May to October. For comments - https://github.com/forestgeo/growth-precip-thailand/issues/10](display/Fig1_new.png)

@@

We identified drought years of interest in the dendroband timeseries - 2010 and 2015 - based on two drought indices and corroborated by meteorological data, local expertise, and plot-level growth responses. 
Within the dendroband timeseries, two ENSO events had occurred; 2010 was a moderate ENSO event while 2015 was a very strong ENSO event (Oceanic Niño Index @bamston_documentation_1997). 
<!-- We chose to identify the years starting from the timeseries, instead of directly using climate variables and their deviation because drought definitions vary widely, and drought characterisation of a year can vary based on the definition selected. -->
We also calculate Standardised Precipitation-Evapotranspiration Index (SPEI) at 1-, 3-, 6- and 12- month scales using the SPEIbase dataset [@vicente-serrano_multiscalar_2010].
Defining drought as periods with two or more consecutive months with 1-month SPEI < -1 also resulted in identification of 2010 and 2015 as drought events. 
These years also corresponded with expert knowledge of major droughts that affected the region as well as meteorological records (see next paragraph; Fig. `r fig_droughts`) and plot-level growth declines in the timeseries (Fig. `r fig_drought_responses`). 
<!-- We report monthly SPEI for the two drought years along with long-term mean SPEI for each month.  -->
<!-- Moreover, broad definitions based on climate variables (often not measured at the location), may not capture ecological realities within the plot.  -->
We note that although tree growth was unusally low in 2020, this year did not emerge as a drought year either in the examination of drought indicators or from expert knowledge.

@@

<!----Contributions beyond---->
Despite consistent observations of growth declines at the plot level across drought events, we show that this scale obscures a variety of ecological responses at the species and tree level that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the two events; it's growth nearly doubled in the dry-season drought but was severely reduced in the wet-season drought (Fig. S5).
More generally, tree-ring records are likely to give a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to directly and interactively influence drought sensitivity.<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in the two different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to drought characteristics suggests potential spatiotemporal insurance effects among tropical tree response to drought, in alignment with recent global analyses (@liu_drought_2022).
